Semi-Permeable Vesicles: Mathematical Formulation, Numerics, and Physics

### 2020-08-18 vesicle_code/docs/permeableVes/main.tex
small changs to paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Examples}
We consider a single semi-permeable vesicle in four different flow
regimes. All vesicles have a length of $L=4\pi$, but the area is not
conserved because of semi-permeability. However, the area is bounded by
$4\pi$ which is the area of a circle with length $L=4\pi$. For each
simulation, we apply an SDC second-order adaptive time stepping
method~\cite{qua-bir2016}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Quiescent Flow} 
We consider a semi-permeable elliptical vesicle with initial reduced
area $\nu = 0.65$ and several different permeability rates in a
quiescent flow. The steady-state of an impermeable vesicle has constant
negative tension~\cite{kra-win-sei-lip1996}, but based
on~\eqref{eqn:SSshape}, a semi-permeable vesicle relaxes to a circle
with positive constant tension $\Lambda = 2\pi^2/L^2$. Recalling the
governing equation for the area in~\eqref{eqn:areaROC}, we expect both
the time for the vesicle to deviate from its initial area, and the time
to reach its steady state will scale with $\beta^{-1}$. In
Table~\ref{tbl:ellipseRelaxTimes}, we report the time,
$t_\mathrm{steady}$, required for the vesicle's area to be within 1.3\%
of its maximum area, and the time, $t_\mathrm{inflate}$, required for
the vesicle's area to exceed 1.3\% of its initial area. We see that both
times scale with $\beta^{-1}$. We also plot the area of a vesicle with
six different permeability rates in Figure~\ref{fig:ellipseArea}. The
area is computed directly from the vesicle shape (solid lines) and by
integrating~\eqref{eqn:areaROC} (marks) with a first-order
method. We also include black dashed lines indicating the times,
$t_{\mathrm{steady}}$, that the vesicle is near its steady state
circular shape.
\begin{table}[htp]
  \centering
  \caption{\label{tbl:ellipseRelaxTimes} The time for a semi-permeable
  vesicle to its steady state ($t_\mathrm{steady}$) and to begin
  inflating ($t_\mathrm{inflate}$).}
  \begin{tabular}{C{0.8cm}C{2cm}C{0.8cm}C{2cm}C{0.8cm}}
    $\beta$ & $t_\mathrm{steady}$ & ratio & $t_\mathrm{inflate}$ & ratio \\
    \midrule
    $10^{0}$  & $6.45 \times 10^{0}$ & ---  
              & $2.17 \times 10^{-2}$ & --- \\ 
    $10^{-1}$ & $2.00 \times 10^{1}$ & 3.10 & 
                $1.25 \times 10^{-1}$ & 5.76 \\
    $10^{-2}$ & $9.61 \times 10^{1}$ & 4.81 & 
                $1.59 \times 10^{0}$ & 12.72 \\
    $10^{-3}$ & $8.36 \times 10^{2}$ & 8.71 & 
                $2.19 \times 10^{1}$ & 13.77 \\
    $10^{-4}$ & $8.15 \times 10^{3}$ & 9.74 & 
                $2.14 \times 10^{2}$ & 9.77 \\
    $10^{-5}$ & $8.03 \times 10^{4}$ & 9.85 & 
                $2.20 \times 10^{3}$ & 10.14 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

We further analyze the transient dynamics for small $\beta$ by
considering the vesicle velocity~\eqref{eqn:vesVelocity} due to
semi-permeability, $\beta(\ff \cdot \nn)\nn$, and force balance,
$\SS[\ff]$. Initially, since $\beta$ is small, the semi-permeability
velocity is much smaller than the force balance velocity. Therefore, the
velocity resembles that of an impermeable vesicle. As the
vesicle approaches the steady state shape of an impermeable vesicle, the
force balance velocity decreases below the semi-permeability velocity,
and then the semi-permeable vesicle begins to dominate, causing the
vesicle to inflate.  As the vesicle becomes more circular, the
membrane force and the corresponding velocity both converge to
zero. In Figure~\ref{fig:vesVelocity} we plot the force balance velocity
(blue) and the semi-permeability velocity (red) at three times with
$\beta=10^{-3}$. We also plot the velocity of an identically initialized
impermeable vesicle (green).  At $t=0.5$, semi-permeability velocity is
small relative to the force balance velocity, and the semi-permeable
velocity matches with the impermeable vesicle velocity. At time $t=5$,
the magnitude of the velocity of the impermeable vesicle and the force
balance velocity are both comparable to the magnitude of the
semi-permeable velocity, and the semi-permeable vesicle begins to
inflate. Then, at time $t=50$, the impermeable vesicle is at its steady
state, while the semi-permeable vesicle still has contributions due to
semi-permeability and force balance. The bottom right plot is the
maximum norm of the different velocity contributions as a function of
time, and the black vertical lines correspond to the times reported in
other three subplots.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Examples}
We consider a single semi-permeable vesicle in quiescent, shear, and
parabolic flows. We also consider a semi-permeable vesicle passing
through a constriction. For each example, all vesicles have a length
$L$, but the area is not conserved because of semi-permeability, but it
is bounded by $L^2/4\pi$ which is the area of a circle with
circumference $L$. For each simulation, we apply an SDC second-order
adaptive time stepping method~\cite{qua-bir2016}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Quiescent Flow} 
We consider a semi-permeable elliptical vesicle with initial reduced
area $\nu = 0.65$ and several different permeability rates in a
quiescent flow. The steady-state of an impermeable vesicle has constant
negative tension~\cite{kra-win-sei-lip1996}, but based
on~\eqref{eqn:SSshape}, a semi-permeable vesicle relaxes to a circle
with positive constant tension $\Lambda = 2\pi^2/L^2$. Since the
governing equation for the area in~\eqref{eqn:areaROC} is proportional
to $\beta$, we expect both the time for the vesicle to deviate from its
initial area, and the time to reach its steady state to be proportional
to $\beta^{-1}$. In Table~\ref{tbl:ellipseRelaxTimes}, we report the
time, $t_\mathrm{steady}$, required for the vesicle's area to be within
1.3\% of its maximum area, and the time, $t_\mathrm{inflate}$, required
for the vesicle's area to exceed 1.3\% of its initial area. Both times
are proportional to $\beta^{-1}$. Next we plot the area of a vesicle
with six different permeability rates in Figure~\ref{fig:ellipseArea}.
The area is computed directly from the vesicle shape (solid lines) and
by integrating~\eqref{eqn:areaROC} (marks) with a first-order method. We
include black dashed lines indicating the times, $t_{\mathrm{steady}}$,
that the vesicle is near its steady state circular shape.
\begin{table}[htp]
  \centering
  \caption{\label{tbl:ellipseRelaxTimes} The time for a semi-permeable
  vesicle to its steady state ($t_\mathrm{steady}$) and to begin
  inflating ($t_\mathrm{inflate}$).}
  \begin{tabular}{C{0.8cm}C{2cm}C{0.8cm}C{2cm}C{0.8cm}}
    $\beta$ & $t_\mathrm{steady}$ & ratio & $t_\mathrm{inflate}$ & ratio \\
    \midrule
    $10^{0}$  & $6.45 \times 10^{0}$ & ---  
              & $2.17 \times 10^{-2}$ & --- \\ 
    $10^{-1}$ & $2.00 \times 10^{1}$ & 3.10 & 
                $1.25 \times 10^{-1}$ & 5.76 \\
    $10^{-2}$ & $9.61 \times 10^{1}$ & 4.81 & 
                $1.59 \times 10^{0}$ & 12.72 \\
    $10^{-3}$ & $8.36 \times 10^{2}$ & 8.71 & 
                $2.19 \times 10^{1}$ & 13.77 \\
    $10^{-4}$ & $8.15 \times 10^{3}$ & 9.74 & 
                $2.14 \times 10^{2}$ & 9.77 \\
    $10^{-5}$ & $8.03 \times 10^{4}$ & 9.85 & 
                $2.20 \times 10^{3}$ & 10.14 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

We further analyze the transient dynamics for small $\beta$ by
considering the vesicle velocity~\eqref{eqn:vesVelocity} due to
semi-permeability, $\beta(\ff \cdot \nn)\nn$, and force balance,
$\SS[\ff]$. Initially, since $\beta$ is small, the semi-permeability
velocity is much smaller than the force balance velocity and the
velocity resembles an impermeable vesicle. As the
vesicle approaches the steady state shape of an impermeable vesicle, the
force balance velocity decreases below the semi-permeability velocity,
and then the semi-permeable velocity begins to dominate, causing the
vesicle to inflate. As the vesicle becomes circular, the
membrane force and the corresponding velocity both converge to
zero. In Figure~\ref{fig:vesVelocity} we plot the force balance velocity
(blue) and the semi-permeability velocity (red) at three times with
$\beta=10^{-3}$. We also plot the velocity of an identically initialized
impermeable vesicle (green). At $t=0.5$, the semi-permeability velocity is
small relative to the force balance velocity, and the semi-permeable
velocity matches with the impermeable vesicle velocity. At time $t=5$,
the magnitudes of the velocity of the impermeable vesicle and the force
balance velocity are comparable to the magnitude of the
semi-permeable velocity, and the semi-permeable vesicle begins to
inflate. Then, at time $t=50$, the impermeable vesicle is at its steady
state, while the semi-permeable vesicle is still tending to a circular
vesicle. The bottom right plot of Figure~\ref{fig:vesVeocity} is the
maximum norm of the different velocity contributions as a function of
time, and the black vertical lines correspond to the times reported in
other three subplots.
